The Gender Pay Gap in the General Social Survey

### 2023-08-23 Paper/paper.Rmd
minor changes in the paper

The prediction of individuals' wages has been an issue in previous literature. As wages are relevant to companies they want to know how predict wages [@Chakraborti]. Some papers make use of classification techniques to predict wages [@Chakraborti]. As we have numeric values of real wages and not a categeorical variable we do not make use of classification techniques but prefer linear regression techniques such as OLS and Lasso regressions as well as random forests. We prefer these techniques as they handle categorical variables and outliers well, which are features of our data [@Cutler]. Wages were already predicted using random forests by [@Eichinger]. They find that usnig random forests results in better predictions than using linear models. Additionally, they show that gender barely effects the wage predictions.

Based on the previous literature, we use severall regression techniques to predict `realrinc`:

@@

In order to evaluate which of the regression techniques predicts `realrinc` the best we split the data set into a train (70%) and a test (30%) dataset. We fit the aforementioned regression models on the train data and evaluate the root mean squared error (RMSE) using the validation data. We choose the RMSE as it is more sensitive to large errors [@Arour].

@@

# Data Preparation

Before conducting the regression analysis we prepared the data. The provided dataset has 54,842 observations and for 38.55% of the observations the outcome variable `realrinc` is missing. As we cannot predict the wage for those observations we removed them. This left us with a dataset containing 33,702 observations. Additionally, we do not know the occupation for some observations (`occ10 = 9997` ) and removed those. This diminishes the dataset to 33,244 observations.

In this dataset still there were missing some values for some observations, e.g., number of children or occupation. To impute these values we used the `mice` function. We imputed age, occupational prestige and number of children with predictive mean matching. For the factor variables education and marital status we used polytomous logisitic regression. We receive six imputation datasets and use mean estimation for age, occupational prestige and number of children.

With this dataset we create dummy variables for the factor variables. Additionally, we create interaction variables.

[Something about ANOVA test for interaction variables].

[Should we consider outlier handling? cf. Eichinger 2022]

In a final step we calculate the correlation matrix using Spearman's rank correlation coefficient. We remove all variables that have a higher than 90% correlation with each other. Our final dataset then contains 33,244 observations of 361 variables.

[Something about how we define the x variables for the models]

@@

We used five different regression techniques to predict the individuals' wages using the final dataset. We fit every model on our training data using the `train` function from the `caret` package. To choose the best parameter combination we use cross-validation within all train functions.

@@

``` r
ridge <- train(formula, data = train,
  method = "glmnet", trControl = trainControl(method = "cv"),
  tuneGrid = expand.grid(alpha = 0, lambda = seq(0, 500, 1))) 
```

>>>>

The prediction of individuals' wages has been an issue in previous literature. As wages are relevant to companies they want to know how predict wages [@Chakraborti]. Some papers make use of classification techniques to predict wages [@Chakraborti]. As we have numeric values of real wages and not a categorical variable we do not make use of classification techniques but prefer linear regression techniques such as OLS and Lasso regressions as well as random forests. We prefer these techniques as they handle categorical variables and outliers well, which are features of our data [@Cutler]. Wages were already predicted using random forests by [@Eichinger]. They find that using random forests results in better predictions than using linear models. Additionally, they show that gender barely effects the wage predictions.

Based on the previous literature, we use several regression techniques to predict `realrinc`:

@@

In order to evaluate which of the regression techniques predicts `realrinc` the best we split the data set into a train (70%) and a test (30%) data set. We fit the aforementioned regression models on the train data and evaluate the root mean squared error (RMSE) using the validation data. We choose the RMSE as it is more sensitive to large errors [@Arour].

@@

# Data Preparation & Feature Engineering

The initial data set contains 11 variables, which are described in the following:

-   **year**: survey year
-   **realrinc**: respondent's base income (in constant 1986 USD)
-   **age**: respondent's age in years
-   **occ10**: respondent's occupation code
-   **occrecode**: recode of the occupation code into one of 11 main categories
-   **prestg10**: respondent's occupational prestige score
-   **childs**: number of children
-   **wrkstat**: work status of the respondent (full-time, part-time, temporarily not working, unemployed (laid off), etc.)
-   **gender**: respondent's gender
-   **educcat**: respondent's degree level (Less Than High School, High School, etc.)
-   **maritalcat**: respondent's marital status (Married, Widowed, Divorced, etc.)

## Data Preparation

Before modeling we conduct a data preparation process. The initial data set has 54,842 observations and the outcome variable `realrinc` has 38.55% of missing values. The imputation of the target variable is discarded to avoid generating a bias in the modeling and therefore we decided to eliminate these records from the original data set. After this, the dataset contains 33,702 observations. Furthermore, it is impossible to deduce the occupation for some observations (`occ10 = 9997`) and we remove these from the data set. At the end, the final data set has 33,244 observations, and we use it for the modeling process.

This data set still had some missing values for the regressors, e.g., number of children, age, marital status, degree level, and others. We use the `mice` package with five variables for the imputation. The features *age, occupational prestige and number of children* are imputed using the pmm (`predictive mean matching`) method. We use the polytomous logistic regression method for the factor variables (*education and marital status*). The imputation process generates 6 different data sets. We apply the estimation of the mean and the majority vote to aggregate the data sets into a single data set. The mean estimation is applied for the numeric variables and majority vote for the categorical variables..

## Feature Engineering

After the imputation process, we create the interaction variables. First, we estimate some interaction features using only the numerical variables. Then, we transform the numerical variables into categorical variables by dividing them into different ranges (e.g., age between 18 and 30, age between 31 and 50, etc.). We transform these new categorical variables and the initial categorical variables (*marital status, education, etc.*) into dummy variables. And finally, the interaction variables are estimated with the group of dummy variables mentioned above. The inclusion of the interaction of the numeric variables are validate using a linear model, and the results show that the p- values of the parameters for the interactions are less than 0.05.

![table](includes/reg_table.png){width="317"}

Furthermore, we estimate a correlation matrix using Spearman's rank correlation coefficient with some numeric variables in order to support the previous results. The correlation results are shown in the following figure.

![matrix_corr](includes/corr_matrix.png)

After the data preparation and feature engineering processes, we construct a final data set with 33,244 observations and 421 features. Due to the large number of interaction variables, we decided to reduce them and use only the interaction between a group of variables (*occrecode, educcat and prestg10*). After reduction, we develop the modeling process with 100 variables (numerical and interaction variables). The following figure shows the overview of the group of variables that were estimated and used in the models.

![feature_process](includes/feature_process.png)

@@

We use five different regression techniques to predict the individuals' wages using the final data set. We fit every model on our training data using the `train` function from the `caret` package. To choose the best parameter combination we use cross-validation within all train functions.

@@

``` r
ridge <- train(formula, data = train,
  method = "glmnet", trControl = trainControl(method = "cv", number = 3),
  tuneGrid = expand.grid(alpha = 0, lambda = seq(0, 500, 1))) 
```
